---
title: "Assignment 3: Profit with Almond Yield"
author: "Carroll and Morris"
date: "2025-04-21"
format: html
code-fold: show
toc: true
editor: visual
embed-resources: true
---
## Load Packages and Functions
```{r, warning = FALSE, message = FALSE}
# load libraries
library(here)
library(tidyverse)
```

```{r}
# call almond yield function
source(here('assignments/assignment3/R/I_Morris-Sibaja_S_Carroll_almond_profit_model.R'))
```
## Apply the Function
```{r}
# read in climate data
clim_data <- read.table(here::here("assignments/assignment3/data/clim.txt"), header = TRUE)
# assign almond parameters to variables
almond_params <- list(
  c_T = -0.015,
  c_T2 = -0.0046,
  c_P = -0.07,
  c_P2 = 0.0043,
  intercept = 0.28
)
# Run function and save df to environment
almond_variables_df <- almond_model(clim_data, almond_params)
```
## Profit Model
```{r}
profit_model <- function(yield_anomaly_df, profit_params) {
  
  profit_df <- yield_anomaly_df %>%
    
    # convert yield anomaly to yield
    mutate(
      yield = profit_params$baseline_yield +  almond_variables_df$yield_anomaly,
      # calculate revenue
      revenue = yield * profit_params$price_per_ton,
      # calculate costs
      labor_costs = profit_params$labor_costs,
      water_costs = almond_variables_df$precipitation_sum_jan * profit_params$water_rate,
      # Calculate profit
      profit = revenue - labor_costs - water_costs,
    )
  return(profit_df)
}
```

## Profit Model Parameters
```{r}
profit_params <- list(
  baseline_yield = 1000, # tons/acre
  price_per_ton = 5000, # $/ton
  labor_costs = 3000, # $/acre
  base_year = 2000,
  water_rate = 2 # dollar cost per inverse mm water 
)
```
## Profit Wrapper
```{r}
almond_yield_profit_model <- function(clim_data, almond_params, profit_params) {
  
  # get yield anomalies
  yield_results <- almond_model(clim_data, almond_params)
  # calculate profits based on yield anomalies
  profit_results <- profit_model(yield_results, profit_params)
  return(profit_results)
}
```

```{r}
almond_yield_profit_model(clim_data, almond_params, profit_params)
```
## Informal Sensitivity Analysis
```{r}
# Profit model function that builds on yield anomaly model
profit_model <- function(yield_anomaly_df, price_per_ton, labor_costs_per_acre, water_rate, baseline_yield, base_year) {
  # Calculate profit
  profit_df <- yield_anomaly_df %>%
    mutate(
      # Calculate actual yield
      actual_yield = baseline_yield + yield_anomaly,
      # Calculate revenue
      revenue = actual_yield * price_per_ton,
      # Calculate costs
      labor_costs = labor_costs_per_acre,
      water_costs = water_rate / (precipitation_sum_jan + 1),
      # Calculate profit before adjustment
      profit_per_acre = revenue - labor_costs - water_costs,
      # Estimate inflation
      year_factor = 1 + 0.02 * (year - base_year),  # 2% change per year
      # Adjust profit based on the year
      adjusted_profit = profit_per_acre * year_factor
    )
  return(profit_df)
}

# Wrapper function that combines yield and profit models
almond_yield_profit_model <- function(clim_data, almond_params, price_per_ton, labor_costs_per_acre, water_rate, baseline_yield, base_year) {
  # Get yield anomalies
  yield_anomaly <- almond_model(clim_data, almond_params)
  
  # Calculate profits
  profit_results <- profit_model(
    yield_anomaly, 
    price_per_ton, 
    labor_costs_per_acre, 
    water_rate, 
    baseline_yield, 
    base_year
  )
  
  return(profit_results)
}

# Informal Sensitivity Analysis
sensitivity_analysis <- function(clim_data, almond_params, n_samples = 100) {
  # Turn off scientific notation
  options(scipen = 999)
  
  # Parameter 1: Price per ton
  deviation_price <- 0.15  # 15%
  base_price <- 5000
  sample_price <- runif(
    n = n_samples,
    min = base_price - (deviation_price * base_price),
    max = base_price + (deviation_price * base_price)
  )
  
  # Parameter 2: Labor cost
  deviation_labor <- 0.15 # 15%
  base_labor_costs <- 3000
  sample_labor <- runif(
    n = n_samples,
    min = base_labor_costs - (deviation_labor * base_labor_costs),
    max = base_labor_costs + (deviation_labor * base_labor_costs)
  )
  
  # Parameter 3: Base year
  sample_base_year <- sample(2000:2020, n_samples, replace = TRUE)
  
  # Fixed parameters
  water_rate <- 2  # $, cost multiplier per inverse mm precipitation
  baseline_yield <- 1000  # tons/acre
  
  # Put samples together
  params <- tibble(
    price_per_ton = sample_price, 
    labor_costs = sample_labor,
    base_year = sample_base_year
  )
  
  # Run model for each parameter combination
  profit_model_results_table <- params %>%
    pmap(function(price_per_ton, labor_costs, base_year) {
      almond_yield_profit_model(
        clim_data = clim_data,
        almond_params = almond_params,
        price_per_ton = price_per_ton,
        labor_costs_per_acre = labor_costs,
        water_rate = water_rate,
        baseline_yield = baseline_yield,
        base_year = base_year
      )
    })
  
  # Extract and combine all results
  years_per_run <- length(unique(profit_model_results_table[[1]]$year))
  
  sample_series <- map_df(profit_model_results_table, ~.x) %>%
    # Add each parameter for every year
    bind_cols(
      params[rep(1:nrow(params), each = years_per_run), ]
    )
    
  return(sample_series)
}

# Create plots
create_sensitivity_plots <- function(sample_series) {
  # Plot 1: Uncertainty in profits through time
  plot1 <- ggplot(sample_series, aes(as.factor(year), adjusted_profit)) +
    geom_boxplot() +
    scale_x_discrete(breaks = function(x) x[seq(1, length(x), 2)]) +
    scale_y_continuous(labels = scales::label_dollar()) +
    labs(x = "Year",
         y = "Adjusted Profit (USD/acre)",
         title = "Uncertainty in Adjusted Almond Profits over Time") +
    theme_minimal()
  
  # Plot 2: How profit varies with base year parameter
  plot2 <- ggplot(sample_series, aes(as.factor(base_year), adjusted_profit)) +
    geom_boxplot() +
    scale_y_continuous(labels = scales::label_dollar()) +
    labs(x = "Base Year (Parameter)",
         y = "Adjusted Profit (USD/acre)",
         title = "Uncertainty in Adjusted Profits by Variation in Base Year") +
    theme_minimal()
  
  return(list(sample_series_plot = plot1, base_year_plot = plot2))
}

# Function to complete the full analysis 
full_analysis <- function(clim_data, almond_params) {
  # Run sensitivity analysis
  sample_series <- sensitivity_analysis(clim_data, almond_params)
  
  # Create plots
  plots <- create_sensitivity_plots(sample_series)
  
  # Return both the data and plots
  return(list(
    data = sample_series,
    plots = plots
  ))
}
```

```{r}
# run the analysis function
full_analysis(clim_data, almond_params)
```

Our profit model results indicate that as variation in base years increase, profit generally decreases. We made the assumption that the profit of almonds is dependent when the value of precipitation and yield anomalies affect costs and revenue. When we ran a sensitivity analysis test by adjusting our parameters 'price_per_ton' and 'labor_costs', we found that the profit per ton varied by +-15% from our baseline profit of $1000 per ton. We made a data frame and populated it with 100 samples generated for each parameter as well as the sample values calculated for adjusted profits. Our function was then applied to each row of our df with `pmap()`, calculating the estimated profit given specific parameter values.

Our first plot indicates that the uncertainty in profits is highest in the years 1995 and 2005-2007. We expect this is due to the yield anomalies and precipitation patterns during these periods. As expected, years with a higher yield anomaly return a larger uncertainty. Overall, variation in uncertainty is low, which may suggest that there are a limited number of years that are statistical outliers.

----

Our profit model results indicate that as XXXXXXXXXXX increase, profit XXXXXXXXXXXX. We made the assumption that the profit of almonds is XXXXXX when the value of XXXXX is XXXXX. When we ran a  sensitivity analysis test by adjusting our parameters 'XXXXXX' and 'XXXXX', we found that the profit per ton varied by +-XXXXXXXXXX % from our baseline profit of $1000 per ton. We made a data frame and populated it with 100 samples generated for each parameter as well as the sample values calculated for XXXXXXX. Our functoin was then applied to each row of our df with `pmap()`, calculating the estimated profit given specific parameter values.
Our first plot indicates that the uncertainty in profits is highest in the years XXXX and XXXX. We expect this is due to the XXXXXXXXXXXXXX and XXXXXXXX. As expected, years with a higher yiled anaomaly return a larger uncertainty Overall, variation in uncertainty is low, which may suggest that there are a limited number of years that are statistical outliers.