---
title: "ESM 230 Assignment 4"
author: IMS
format: html
code-fold: show
toc: true
editor: visual
embed-resources: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(sensitivity)
library(tidyverse)
library(gridExtra)
library(purrr)
library(ggpubr)
library(here)
```

## Load the atmospheric conductance model function

```{r load_model}
# Load the Catm function from external R script
source(here("assignments", "assignment4", "Catm.R"))
```

## Generate parameter samples

```{r generate_parameters}
# Number of parameter samples
num_samples <- 1000

# Parameter distributions:
# k_o: stomatal conductance coefficient (mean = 0.1, SD = 1% of mean)
k_o_samples <- rnorm(n = num_samples, 
                     mean = 0.1, 
                     sd = 0.1 * 0.01)

# k_d: boundary layer conductance coefficient (mean = 0.7, SD = 1% of mean)
k_d_samples <- rnorm(n = num_samples, 
                     mean = 0.7, 
                     sd = 0.7 * 0.01)

# v: windspeed (mean = 300 cm/s, SD = 50 cm/s)
wind_speed_samples <- rnorm(n = num_samples, 
                            mean = 300, 
                            sd = 50)

# height: vegetation height (uniform between 3.5 and 5.5 meters)
veg_height_samples <- runif(n = num_samples, 
                            min = 3.5, 
                            max = 5.5)

# Combine into first sample matrix (X1)
param_set_X1 <- data.frame(k_o = k_o_samples,
                           k_d = k_d_samples,
                           v = wind_speed_samples,
                           height = veg_height_samples)

# Generate second independent sample matrix (X2)
k_o_samples_2 <- rnorm(n = num_samples, 
                       mean = 0.1, 
                       sd = 0.1 * 0.01)
k_d_samples_2 <- rnorm(n = num_samples,
                       mean = 0.7, 
                       sd = 0.7 * 0.01)
wind_speed_samples_2 <- rnorm(n = num_samples, 
                              mean = 300, 
                              sd = 50)
veg_height_samples_2 <- runif(n = num_samples, 
                              min = 3.5, 
                              max = 5.5)

param_set_X2 <- data.frame(k_o = k_o_samples_2,
                           k_d = k_d_samples_2,
                           v = wind_speed_samples_2,
                           height = veg_height_samples_2)
```

## Set up Sobol sensitivity analysis

```{r setup_sobol}
# Create Sobol object using Saltelli sampling
sobol_results <- sobolSalt(model = NULL, 
                           X1 = param_set_X1, 
                           X2 = param_set_X2, 
                           nboot = 100)
```

## Run the model across sampled parameters

```{r run_model}
# Extract parameter combinations used by Sobol design
params <- as.data.frame(sobol_results$X)
colnames(params) <- colnames(param_set_X1)

# Evaluate the model (Catm function) for each parameter set
conductance_results <- pmap_dbl(params, Catm)

# Attach results back to Sobol object
sens_Catm_Sobol <- tell(sobol_results, conductance_results, res.names = "ga")
```

## Plot distribution of conductance estimates

```{r plot_histogram}
# Combine parameter matrix with model outputs
result_data <- cbind(params, conductance = sens_Catm_Sobol$y)

# Plot histogram with mean conductance line
hist_plot <- ggplot(result_data, aes(x = conductance)) +
  geom_histogram(bins = 30, fill = "red", color = "black") +
  geom_vline(xintercept = mean(result_data$conductance), color = "black", linetype = "dashed") +
  labs(title = "Conductance Estimates Distribution",
       x = "Conductance", y = "Frequency")

print(hist_plot)
```

## Find second most important paramter
```{r}
both <- cbind.data.frame(params, gs = sens_Catm_Sobol$y)

# look at response of conductance and wind speed with height (second in total effect)
v_gs_plot <- ggplot(both, aes(v, gs, col = height)) +
  geom_point() +
  labs(y = "Conductance (mm/s)", x = "Windspeed")

# look at response of conductance and height with wind speed
h_gs_plot <- ggplot(both, aes(height, gs, col = v)) +
  geom_point() +
  labs(y = "Conductance (mm/s)", x = "Vegetation Height")

print(v_gs_plot)
print(h_gs_plot)
```

Estimate the Sobol Indices for your output
```{r}
# main effect:  partitions variance (main effect without co-variance) - sums approximately to one
sens_Catm_Sobol$S

# useful to add names
row.names(sens_Catm_Sobol$S) <- colnames(params)
sens_Catm_Sobol$S

# total effect - accounts for parameter interactions
row.names(sens_Catm_Sobol$T) <- colnames(params)
sens_Catm_Sobol$T

# Both the main effect and total effect can tell us something about how the parameter influences results


print(sens_Catm_Sobol)
```

## Reflection on parameter sensitivity

The plot of wind speed and conductance had a stronger correlation than the in class example with lower and less variable windspeed and taller vegetation. We can conclude from our model that  atmospheric conductance is indeed sensitive to variations in wind speed, as conductance increases consistently with increasing wind speed. Additionally, when vegetation is shorter, we see that its importance increases compared to our class example, reminding us how variables confound and change our model results depending on the system we are wokring with.
